\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{geometry}
\geometry{margin=1in}

\title{Mini Game Hub: Project Report}
\author{Daksh Maahor (25B0974) and Rujul Garg (25B0937)}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
This report details the implementation of Mini Game Hub, a two-player game launcher built with Python and Pygame. It covers the external libraries used, features implemented, code logic explanations, challenges faced, future improvements, and references.
\end{abstract}

\section{External Libraries Used}

The project utilizes the following external libraries:

\begin{itemize}
    \item \textbf{Pygame}: A cross-platform set of Python modules designed for writing video games. Used for rendering the graphical user interface, handling user input, and managing the game loop.
    \item \textbf{NumPy}: A library for the Python programming language, adding support for large, multi-dimensional arrays and matrices, along with a large collection of high-level mathematical functions. Used for efficient board state management and win condition checking in games.
    \item \textbf{Matplotlib}: A plotting library for Python and its numerical mathematics extension NumPy. Used to generate statistical charts and visualizations for the leaderboard data.
\end{itemize}

\section{Features Implemented}

The Mini Game Hub includes the following features:

\begin{itemize}
    \item \textbf{Authentication System}: Secure user login and registration using SHA-256 hashed passwords stored in a TSV file.
    \item \textbf{Game Menu}: A graphical menu allowing users to select from three games: Connect Four, Othello, and Tic-Tac-Toe.
    \item \textbf{Game Implementations}: Full implementations of three board games with proper win detection, turn management, and visual feedback.
    \item \textbf{Leaderboard}: A command-line script that processes game history and displays sorted statistics.
    \item \textbf{Statistics Dashboard}: Matplotlib-generated charts showing player performance and game popularity.
    \item \textbf{How-to-Play Guide}: Interactive in-game instructions for each game mode.
    \item \textbf{Persistent Data Storage}: Game results saved to CSV files for later analysis.
\end{itemize}

\section{Code Logic Explanation}

\subsection{Generic Class (BoardGame)}

The \texttt{BoardGame} class serves as a base class for all board games, providing common functionality:

\begin{itemize}
    \item \textbf{Initialization}: Sets up the board as a NumPy array, player names, and turn management.
    \item \textbf{Turn Management}: Alternates between players using a simple index toggle.
    \item \textbf{Abstract Methods}: Defines \texttt{check\_win()} as an abstract method that subclasses must implement.
\end{itemize}

\subsection{Game Implementations}

\subsubsection{Connect Four}
\begin{itemize}
    \item Uses a 7x7 grid where pieces fall to the lowest available row in a column.
    \item Win detection checks for four consecutive pieces horizontally, vertically, or diagonally using NumPy array slicing.
    \item Rendering includes hover previews and animated piece drops.
\end{itemize}

\subsubsection{Othello (Reversi)}
\begin{itemize}
    \item Played on an 8x8 board with initial four pieces in the center.
    \item Move validation uses vectorized NumPy operations to check for valid captures in all eight directions.
    \item Disc flipping is implemented by iterating through captured sequences and updating the board.
\end{itemize}

\subsubsection{Tic-Tac-Toe (Expanded)}
\begin{itemize}
    \item Uses a 10x10 grid requiring five in a row to win.
    \item Win detection employs sliding window views to efficiently check all directions for winning sequences.
    \item Includes hover hints and visual highlighting of winning cells.
\end{itemize}

\subsection{Authentication System}

The authentication system is implemented in Bash:

\begin{itemize}
    \item \textbf{User Input Validation}: Strips control characters and rejects invalid characters to prevent injection attacks.
    \item \textbf{Password Hashing}: Uses SHA-256 to hash usernames and passwords before storage.
    \item \textbf{File Storage}: Stores hashed credentials in a tab-separated values file.
    \item \textbf{Login Flow}: Hashes input and compares against stored hashes for authentication.
\end{itemize}

\subsection{Leaderboard}

The leaderboard is generated by an AWK script:

\begin{itemize}
    \item \textbf{Data Processing}: Parses CSV history files to count wins, losses, and calculate win ratios per player per game.
    \item \textbf{Sorting}: Uses Unix sort to order results by the specified metric (wins, losses, or ratio).
    \item \textbf{Output Formatting}: Prints tabular results grouped by game type.
\end{itemize}

\section{Hurdles Faced and Solutions}

Several challenges were encountered during development:

\begin{itemize}
    \item \textbf{Input Validation}: Initial username input allowed control characters, causing display issues. Solved by implementing character filtering using \texttt{tr} and \texttt{sed} in Bash.
    \item \textbf{Pygame Rendering}: Coordinating multiple UI elements (buttons, text, boards) required careful positioning calculations. Resolved by creating reusable drawing functions and consistent coordinate systems.
    \item \textbf{Win Detection Algorithms}: Implementing efficient win checking for different games. Overcame by leveraging NumPy's array operations for vectorized computations.
    \item \textbf{Cross-Platform Compatibility}: Ensuring the application works on different systems. Addressed by using relative paths and Python's pathlib for file operations.
\end{itemize}

\section{What We Would Do Differently with 10 More Hours}

With additional time, we would implement:

\begin{itemize}
    \item \textbf{Enhanced UI}: Add animations, sound effects, and a more polished visual design.
    \item \textbf{Additional Games}: Implement more board games like Chess or Checkers.
    \item \textbf{Network Play}: Add online multiplayer functionality using sockets or a web framework.
    \item \textbf{Database Integration}: Replace file-based storage with a proper database for better scalability.
    \item \textbf{Unit Tests}: Add comprehensive test suites for game logic and authentication.
    \item \textbf{Configuration System}: Allow users to customize game settings and themes.
\end{itemize}

\section{Bibliography}

\begin{enumerate}
    \item Pygame Documentation. \url{https://www.pygame.org/docs/}. Accessed April 2026.
    \item NumPy User Guide. \url{https://numpy.org/doc/stable/user/index.html}. Accessed April 2026.
    \item Matplotlib Documentation. \url{https://matplotlib.org/stable/contents.html}. Accessed April 2026.
    \item Python Documentation. \url{https://docs.python.org/3/}. Accessed April 2026.
    \item Bash Reference Manual. \url{https://www.gnu.org/software/bash/manual/bash.html}. Accessed April 2026.
\end{enumerate}

\end{document}
